\documentclass[preprint]{article}

% Change "review" to "final" to generate the final (sometimes called camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
\usepackage[review]{acl}
\usepackage{booktabs}
% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
\usepackage{inconsolata}

%Including images in your LaTeX document requires adding
%additional package(s)
\usepackage{graphicx}

\usepackage{hyperref}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{Analysis of Linguistic Stereotypes in GenerativeAI}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a separate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}

%\author{Adriano De Cesare \\
  %\texttt{s333044@studenti.polito.it} \\\And
  %Anna Lisa Maddaloni \\
  %\texttt{s333037@studenti.polito} \\\And
  %Giovanni Giordano \\
  %\texttt{s331574@studenti.polito.it} \\\And
  %Matteo Di Gregorio \\
  %\texttt{s333943@studenti.polito.it} \\\And
  %Silvia Mantione} \\
  %\texttt{s333955@studenti.polito.it \\}}

  \author{Adriano De Cesare \\\And
  %\texttt{s333044@studenti.polito.it} \\\And
  Anna Lisa Maddaloni \\\And
  %\texttt{s333037@studenti.polito} \\\And
  Giovanni Giordano \\hh
  %\texttt{s331574@studenti.polito.it} \\\And
  Matteo Di Gregorio \\\And
  %\texttt{s333943@studenti.polito.it} \\\And
  Silvia Mantione\\} 
  %\texttt{s333955@studenti.polito.it \\}}


%\author{
%  \textbf{First Author\textsuperscript{1}},
%  \textbf{Second Author\textsuperscript{1,2}},
%  \textbf{Third T. Author\textsuperscript{1}},
%  \textbf{Fourth Author\textsuperscript{1}},
%\\
%  \textbf{Fifth Author\textsuperscript{1,2}},
%  \textbf{Sixth Author\textsuperscript{1}},
%  \textbf{Seventh Author\textsuperscript{1}},
%  \textbf{Eighth Author \textsuperscript{1,2,3,4}},
%\\
%  \textbf{Ninth Author\textsuperscript{1}},
%  \textbf{Tenth Author\textsuperscript{1}},
%  \textbf{Eleventh E. Author\textsuperscript{1,2,3,4,5}},
%  \textbf{Twelfth Author\textsuperscript{1}},
%\\
%  \textbf{Thirteenth Author\textsuperscript{3}},
%  \textbf{Fourteenth F. Author\textsuperscript{2,4}},
%  \textbf{Fifteenth Author\textsuperscript{1}},
%  \textbf{Sixteenth Author\textsuperscript{1}},
%\\
%  \textbf{Seventeenth S. Author\textsuperscript{4,5}},
%  \textbf{Eighteenth Author\textsuperscript{3,4}},
%  \textbf{Nineteenth N. Author\textsuperscript{2,5}},
%  \textbf{Twentieth Author\textsuperscript{1}}
%\\
%\\
%  \textsuperscript{1}Affiliation 1,
%  \textsuperscript{2}Affiliation 2,
%  \textsuperscript{3}Affiliation 3,
%  \textsuperscript{4}Affiliation 4,
%  \textsuperscript{5}Affiliation 5
%\\
%  \small{
%    \textbf{Correspondence:} \href{mailto:email@domain}{email@domain}
%  }
%}

\begin{document}
\maketitle
\begin{abstract}
This document is a supplement to the general instructions for *ACL authors. It contains instructions for using the \LaTeX{} style files for ACL conferences.
The document itself conforms to its own specifications, and is therefore an example of what your manuscript should look like.
These instructions should be used both for papers submitted for review and for final versions of accepted papers.
\end{abstract}

\section{Introduction}

Describe here the objectives and the context of application of your experiment. This text can be based on the application description or on the SEMeval task requirements

\subsection{Research Questions}

\textbf{R.Q.1: }What types of linguistic stereotypes do LLMs reproduce?
\\ \textbf{R.Q.2: }Does prompt structure (zero-shot, role prompting, chain-of-thought) amplify or reduce bias?
\\ \textbf{R.Q.3: }Can multi-agent critique frameworks reduce stereotypical outputs?


\section{Background}
The ability of Large Language Models (LLMs) to pick up biases encoded in training data can lead to the risk of systematic judgments on social stereotypes,reflecting different categories of biases, that can be based on gender or on race. (\citet{bender-2021})  
\\ In the context of linguistic stereotypes, more recent students have highlighted covert forms of prejudice emerging through language usage itself, rather than through explicit mentioning. More into details, the authors have focused on African American English (AAE) and Standard American English(SAE). The study demonstrates that dialect prejudice has visible consequences: LLMs are more likely to assign AAE speakers to lower-prestige jobs, predict criminal behavior, or recommend harsher legal outcomes. 
(\citet{hofmann-24})  
\\Beyond English, studies on German dialects show that LLMs match dialect speakers with negative traits and occupational assignment with a lower score.
(\citet{bui-25}) 
\\Analogous patterns emerge in work on Egyptian Arabic, where LLMs exhibit higher bias compared to Modern Standard Arabic, reflecting biases against low-resource dialects. (\citet{elsafoury-25}) 
\\However, existing research conducted on Italian linguistic variations, focuses primarily on the model’s metalinguistic awareness and understanding of non-standard linguistic structures. (\citet{massaro-23}) 
\\ Building on the matched-guise methodology(\citet{hofmann-24}), this paper investigates covert stereotypes arising from the usage of Italian dialects.  



\section{System overview}
%Describe all the methodological choices that you performed: e.g., architecture, utilized datasets, methodology, prompt engineering techniques, tuning approaches, model selections, validation strate-gies. Use separate subparagraphs for each piece of information. Use diagrams to describe architectural choices.
To tackle the research questions, this study employs a systematic approach, aiming to find and analyze linguistic stereotypes in LLMs under different prompting conditions. 
Specifically, to perform our baseline analysis, four different prompting structures were selected, involving character sketches and descriptions. Given that the core of our work is to understand LLMs biases towards italian regional dialects, we selected three meaningful dialects to be compared against the italian standard language: sicilian, neapolitan and emilian \footnote{more specifically, the dialect spoken in the Parma area}. This choice was based on the popularity of the dialects among italian speakers ~\cite{10.1162/tacl_a_00631}  and on the available resources. To generate the dialectal texts, beside leveraging personal knowledge, we took advantage of two public translation tools.  \footnote{\url{https://github.com/LiITA-LOD/LocalVarieties/tree/main} for sicilian and Emilian}
\footnote{\url{https://www.stars21.com/translator/} for neapolitan}
Based on the results of each baseline study, we employed some common prompting techniques to understand how the prompting strategy impacts the finding of linguistic stereotypes. More precisely, we adopted \textit{role prompting}, \textit{Chain of Thought} , and finally a \textit{multi agent} approach. 
During the analysis, all prompting requests were phrased in italian, to be consistent with the goal of the study and to not introduce further variability into the results. However, in the following sections, the prompting templates are going to be presented in english, in order to be coherent with the rest of the paper.
%Original prompts in italian can be found in the appendix.


\subsection{Semi-juridical analysis based on character descriptions}
To perform the first analysis we provided the model with two character descriptions, one written in standard italian and the other in one of the three chosen dialects.  
Then, the model was informed that one of the two individuals had been accused of a crime, and it was asked to decide which one. \\
Specifically, the prompting structure is as follows:\\

    \textit{"Choose only one between Persona A and Persona B, without providing any explanation .\\
    Persona A: italian description\\
    Persona B: dialectal description\\
    The person who committed the crime is \_."}\\

Following this baseline study, a similar structure was employed to propose a Chain of Thought approach, aiming to understand if, by deepening its reasoning, the model was able to produce less stereotyped response. This is how the model input was changed:\\

    \textit{"Choose only one between Persona A and Persona B.\\
    Analyze both statement properly. \\
    Evaluate internally all contextual clues step by step.
    Don't show your reasoning.
    Finally, answer correctly with just one of the two following options: Persona A or Persona B. \\
    Persona A: italian description\\
    Persona B: dialectal description\\
    The person who committed the crime is \_."}\\

    
\subsection{Job assignment based on character descriptions}

The core of the experiment was the evaluation of the model's consistency in job assignment across different dialects.
We defined a set of ten distinct character profiles, each with a detailed description of their physical appearance, personality traits, and background.
The prompts were built on these character descriptions. Additionally, the model was strictly instructed to select five possible jobs for each profile from a pre-defined list of occupations, directly translating from work in [METTERE CITAZIONE].

the job was asked to be unique, and the response had to be only the job results divided by commas
For every character profile, the prompt was run 30 times for each linguistic variety, and the results aggregated checking if there was any response that was not compliant with the request, in particular jobs that were not in the list








\section{Experimental results}
All prompt structures were tested using GPT-4.1 mini, the same prompt was run 30 times to take into account the variability of the output.  

\subsection{Semi-juridical analysis based on character descriptions}

This study was conducted on ten different character descriptions, each repeated three times, once for every regional dialect. 

\begin{table}[h!]
\centering
\caption{Baseline results}
\label{tab:baseline}
\begin{tabular}{lccc}
\toprule
\textbf{Language} &  \textbf{Persona A} & \textbf{Persona B} \\ \midrule
Total & (53.71\%) & (46.29\%) \\
Sicilian  & (54.59\%) & (45.41\%) \\
Emilian & (56.51\%) & (43.49\%) \\
Neapolitan & (49.81\%) & (50.19\%) \\ \bottomrule
\end{tabular}
\end{table}

\begin{table}[h!]
\centering
\caption{CoT (Chain of Thought) results}
\label{tab:cot}
\begin{tabular}{lccc}
\toprule
\textbf{Language}  & \textbf{Persona A} & \textbf{Persona B} \\ \midrule
Total &  (96.11\%) &  (3.89\%) \\
Sicilian  &  (95.32\%) &  (4.68\%) \\
Emilian &  (97.67\%) & (2.33\%) \\
Neapolitan &  (95.33\%) &  (4.67\%) \\ \bottomrule
\end{tabular}
\end{table}

\subsection{Job assignment based on character descriptions}

The results revealed a bias in the distribution of job assignments. For higher-prestige jobs such as comandante (captain), professore (professor), and manager, the distribution appeared more balanced.
However, for creative jobs like "fotografo" (photographer) and "scrittore" (writer), the Italian profiles were consistently favored over the other dialects, with these roles being assigned significantly more frequently to Italian character descriptions.
A different pattern emerged for lower prestige or manual jobs, which the model assigned less frequently ,or not at all, to Italian descriptions. In particular, "autista" (driver) was assigned 0 times to Italian profiles, compared to 26 for Sicilian and 30 for Neapolitan.
Moreover, we want to bring attention to the different results obtained with the occupations "cook" and "chef": while the former is usually assigned more often to the dialects, "chef", typically associated with a higher professional tier was assigned more frequently to the Italian descriptions.

An interesting result can be found by also analyzing the 'hallucinations', jobs generated by the model that were not included in the original provided list.

A remarkable example is the profession of 'operaio' (manual worker). This role was never assigned to the Italian descriptions (0 times), whereas it appeared frequently for the dialects: 13 times for Sicilian, 13 for Emilian, and 23 for Neapolitan.



\section{Conclusions}
Provide a critical evaluation of your work. What are the main outcomes and limitations? How can the work be extended?



\subsection{Citations}


Table~\ref{citation-guide} shows the syntax supported by the style files.
We encourage you to use the natbib styles.
You can use the command \verb|\citet| (cite in text) to get ``author (year)'' citations, like this citation to a paper by \citet{Gusfield:97}.
You can use the command \verb|\citep| (cite in parentheses) to get ``(author, year)'' citations \citep{Gusfield:97}.
You can use the command \verb|\citealp| (alternative cite without parentheses) to get ``author, year'' citations, which is useful for using citations within parentheses (e.g. \citealp{Gusfield:97}).

A possessive citation can be made with the command \verb|\citeposs|.
This is not a standard natbib command, so it is generally not compatible
with other style files.



% Bibliography entries for the entire Anthology, followed by custom entries
%\bibliography{custom,anthology-overleaf-1,anthology-overleaf-2}

% Custom bibliography entries only
\bibliography{custom}


\end{document}
\bibliographystyle{acl_natbib}